Medicare Spending 
========================================================

```{r include=FALSE}
require(DCF)
```

In May 2013, the [Centers for Medicare and Medicaid Services](http://www.cms.gov/Research-Statistics-Data-and-Systems/Statistics-Trends-and-Reports/Medicare-Provider-Charge-Data/index.html) released data on the price of health services at hospitals around the US.  This attracted press attention.  For example, a headline in the New York Times read ["Hospital Billing Varies Wildly, Government Data Shows"](http://www.nytimes.com/2013/05/08/business/hospital-billing-varies-wildly-us-data-shows.html).  A [map with the article](http://www.nytimes.com/interactive/2013/05/08/business/how-much-hospitals-charge.html?_r=0) shows the geographic disparities in hospital charges.  

* Read the New York *Times* article to familiarize yourself with the issue.

The data themselves are available as a [zip-compressed CSV file](http://www.cms.gov/Research-Statistics-Data-and-Systems/Statistics-Trends-and-Reports/Medicare-Provider-Charge-Data/Downloads/IPPS_DRG_CSV.zip).  

DRAFTY: This file format happens to be wasteful in storage and speed for reasons we will discover later in the semester.  You can read in the data frame, called `med` with: (PUT THIS IN THE PACKAGE or on a fetch repository)
```{r}
load("/Users/kaplan/kaplanfiles/Github/DataAndComputingFundamentals/CaseStudies/MedicareSpending/CMS-data-2013-05-06.rda")
```

### Uses for these data

1. How many Direct Recovery Groups are there?
2. What is the most expensive group (on average and in the median)?
3. What is the most expensive state (on average and in the median)?
4. What is the most expensive city (on average and in the median)?
5. What DRG has the largest spread of values? (use the sd)
6. As a proportion from the median, which DRG has the largest spread of values? (use the sd divided by the mean)


Later on (when we do maps): 
1. Find the z-score for each DRG.  Make a map of the average z-score by state.

When we do join:
1. Look at how the data are not normalized.  Present a normalized version of the data and show how much smaller it is.  Link in some other variable relating to zip code, for example per capita income.


