Registration for DCF in Spring 2013
========================================================
```{r include=FALSE}
require(mosaic)
fetchData("survey-processing.R")
require(xtable)
options(xtable.type="html")
```

As of `r I(date())`.

```{r echo=FALSE,message=FALSE}
enroll <- fetchGoogle("https://docs.google.com/spreadsheet/pub?key=0Av2C2RiwUxpVdE1SUElWaVRaNVJkeW4waWQ0QzBoWkE&single=true&gid=0&output=csv")
```

We currently have `r nrow(enroll)` who have signed up on the information sheet for the DCF course for Spring 2013.

### Meeting Time

Students indicated availability for different time slots:

#### Tuesdays 3:00-4:30 pm: `r table(enroll[[7]])[1]` available

#### Tuesdays 4:30-6:00 pm: `r table(enroll[[6]])[1]` available

#### Tuesdays 7:00-8:30 pm: `r table(enroll[[5]])[1]` available

### Interest Areas

```{r echo=FALSE,results="asis"}
fields = c("Bio","Econ","Env","Math","Phys","Poli","Soc","Stat")
for (k in 8:15) {
enroll[[k]] <- factor(enroll[[k]],levels=c("Not at all","Weak interest", "Somewhat", "Strong interest","My main interest"), 
                      labels=c("none","weak","some", "strong","main"),
                      ordered=TRUE)
cat(paste("#### ", fields[k-7],"\n\n<pre>\n"))
print(table(enroll[[k]]))
cat("</pre>\n")
}
```

### Computer Experience

```{r include=FALSE}
for (k in 16:18){
  inds = enroll[[k]]==""
  enroll[[k]][inds] = NA
  enroll[[k]] = factor( as.character(enroll[[k]]))
}
```

#### Spreadsheets

```{r echo=FALSE,results="asis"}
print(table(enroll[[18]]))
```

#### Programming

```{r echo=FALSE,results="asis"}
xtable(table(enroll[[17]]))
```

#### R

```{r echo=FALSE,results="asis"}
print(table(enroll[[16]]))
```