Constructing the Simple Area/Population file
=============================

The raw data are from the UN Food and Agriculture Organization.  In this document, the format is changed to make them a standard table so that students can work with the data effectively from the start.

Use the `reshape2` package
```{r include=FALSE}
require(reshape2)
```

Read in the "raw" data:
```{r}
pop <- read.csv("CountryPopulation.csv")
area <- read.csv("CountryArea.csv")
```

Create a population dataframe that has separate columns for each population variable:
```{r}
goo <- dcast( pop, Country + Year ~ Element, value.var="Value")
```

Similarly, create an area dataframe that has separate columns for each area variable:
```{r}
foo <- dcast(area, Country+Year ~ Item, value.var="Value")
```

Put them into one, wider dataframe and fix the names:
```{r}
FAO <- merge(foo,goo)
nms <- names(FAO)
nms <- sub('[[:space:]]+$', '', nms)
names(FAO) <- nms
```

Write the data, just for 2009:
```{r}
write.csv(file="FAO-simple.csv", subset(FAO,Year==2009), row.names=FALSE)
```