Constructing the Simple Area/Population file
=============================

The raw data are from the UN Food and Agriculture Organization.  In this document, the format is changed to make them a standard table so that students can work with the data effectively from the start.

Use the `reshape2` package
```{r include=FALSE}
require(reshape2)
```

### Population and Area

Read in the "raw" data:
```{r}
pop <- read.csv("CountryPopulation.csv")
area <- read.csv("CountryArea.csv")
```

Create a population dataframe that has separate columns for each population variable:
```{r}
goo <- dcast( pop, Country + Year ~ Element, value.var="Value")
```

Similarly, create an area dataframe that has separate columns for each area variable:
```{r}
foo <- dcast(area, Country+Year ~ Item, value.var="Value")
```

Put them into one, wider dataframe and fix the names:
```{r}
FAO <- merge(foo,goo)
nms <- names(FAO)
nms <- sub('[[:space:]]+$', '', nms)
names(FAO) <- nms
```

Write the data, just for 2009:
```{r}
write.csv(file="FAO-simple.csv", subset(FAO,Year==2009), row.names=FALSE)
```

### Food Supply

Read in the data files.  The raw files, downloaded in bulk mode from the UN FAO, should be opened in Excel and saved as CSV files.  

An example has been saved as `foobar.csv`

```{r}
raw <- read.csv("FAO_DATA/foobar.csv")
#raw2 <- raw[,c(1:9,seq(11,length(raw),by=2))] # get rid of F columns
raw2009 <- raw[,c(1:8,(length(raw)-1))]
```

Delete the `(Total)` columns?