Instructor Notes: Simplifying the Birds Dataset
========================================================

```{r include=FALSE}
require(DCF)
```

The Ordway bird data has lots of duplicate coding of species and many other errors.  Initially, the students are going to work with a simplified version of the data.

```{r}
data(OrdwayBirds)
count <- data.frame(tally( ~species.name, data=OrdwayBirds))
names(count) <- c("species.name","species.count")
OrdwaySimple <- join(OrdwayBirds, count, by="species.name")
# OrdwaySimple <- droplevels(subset(OrdwaySimple, species.count > 200))
OrdwaySimple <- droplevels(OrdwaySimple)
```

Change the Time.of.capture to be just the hour, and clean:
```{r}
bad = OrdwaySimple$Time.of.capture %in% c("M/4","none")
OrdwaySimple$Time.of.capture[bad] = NA
T = strsplit( as.character(OrdwaySimple$Time.of.capture),":",fixed=TRUE)
OrdwaySimple$Hour.of.capture = sapply(T, FUN=function(x){as.numeric(x[[1]])})
bad = !OrdwaySimple$Hour.of.capture %in% 0:24
OrdwaySimple$Hour.of.capture[bad]=NA
```

Convert the potentially numerical values into simple numbers.
```{r}
OrdwaySimple$Weight = as.numeric(as.character(OrdwaySimple$Weight))
OrdwaySimple$Wing.chord = as.numeric(as.character(OrdwaySimple$Wing.chord))
OrdwaySimple$Tail.length = as.numeric(as.character(OrdwaySimple$Tail.length))
```

Write the file.
```{r}
write.csv(OrdwaySimple, "OrdwaySimple.csv", row.names=FALSE)
```

Then post the file for use with `fetchData`.

