Creating Small Data Tables by Hand
=======================

```{r include=FALSE}
require(DCF)
```

It sometimes happens that you have a small set of data that you need to enter into the computer.  This might be because you've done a small study, or because you have some auxiliary data that you need to help with a larger study.  In either event, for entering small data sets, a spreadsheet is an appropriate tool.

We'll use Google Spreadsheets as our tool for data entry.  In addition to being available for free, it's effective for collaborating among several people.

### Publishing a Google Spreadsheet to R.

For each spreadsheet, you need to generate a URL that instructs Google to publish the spreadsheet on demand, so that R can read the spreadsheet.  This is a one-time operation that will continue to work even if you add more data to the spreadsheet.

1. In Google spreadsheets,  choose `File/Publish to the web`
2. Start publishing, checking “Automatically republish when changes are made”
3. Under “Get a link to the published data” Select “CSV” Sheet 1 All cells
4. Copy the link. You'll need it in a bit.

Now, to read the spreadsheet in to R …
1. Paste the link from step (5) into R as a quoted argument to fetchGoogle() command, like this

```{r results="hide",messages=FALSE,warning=FALSE,error=FALSE}
mydat = fetchGoogle("https://docs.google.com/spreadsheet/pub?key=0Am13enSalO74dHJHcjktU3k3NFFXblpMQ25WcDdnMVE&single=true&gid=0&output=csv")
```

<iframe width="420" height="315" src="http://www.youtube.com/embed/8ZA4os9MMac?rel=0" frameborder="0" allowfullscreen></iframe>