Grouping Operations
============

```{r include=FALSE}
require(DCF)
options(na.rm=TRUE)
```

Cases often fall into groups defined by some measured variable.  Often, your interest is in the group rather than the individual case.  There is a grouping variable and an operation to be performed on the group.  Some examples:
* How many birds are there of each species?  The grouping variable is species, the operation is the number of cases.
* What's the average weight of people who are between 1.6 and 1.7 m tall?  The grouping variable is height, the operation is averaging the weight.
* What's the total population of countries smaller than 100 sq. km?  The grouping variable is land area, the operation is summing the population.

In the examples given here, the R operations are `length()` for counting, `mean()` for averaging, and `sum()` for adding.

The R function that we will use for grouping is `groupBy()`, provided by the DCF package.  This takes three arguments:
1. `data=` the data frame on which to perform grouping.
2. `by=` a grouping variable in the data frame or computed from it.
3. The operation to perform.

Statements corresponding to the three examples above.

#### How many birds are there of each species?

```{r}
data(OrdwayBirds)
countBySpecies = groupBy(data=OrdwayBirds, 
                         by=species.name, 
                         scount=length(species.name))
head(countBySpecies)
```
Two important things to note:
1. The use of the `length()` function to count the number of cases
2. Any variable could have been used as the input to `length()`; they all have the same length.

#### Average weight of people between 1.6 and 1.7m tall.

```{r}
data(nhanes)
nhanes = transform(nhanes, 
                   hgroups=cut(hgt,breaks=seq(0.5,2.0,by=.1)))
meanWeight = groupBy(data=nhanes, 
                     by=hgroups, 
                     weight=mean(wgt), height=mean(hgt))
head(meanWeight)
```

Things to note:
1. Since `hgt` is a quantitative variable, we created a new categorical  variable, `hgroups`, that splits up `hgt` into discrete groups.
2. The `meanWeight` data frame produced by `groupBy` has been arranged to label the mean weight variable as `weight`.
3. For later convenience, a quantitative variable `height` has been produced so that weight can be compared to height quantitatively.


#### What's the total population of countries smaller than 100 sq. km?

```{r}
data(FAOsimple)
totalPop = groupBy( data=FAOsimple,
                    by=list(small=Country.area < 100),
                    population=sum(Total.Population...Both.sexes) )
totalPop
```

Things to note ...
* The `by` criterion is being defined through a calculation, rather than just naming a variable. The point of using `list()` is to give a simple name to the result.
* The result of the summing is being stored in a variable called `population` 

### More than one grouping variable

Sometimes you are interested in dividing groups into subgroups.  You can do this by specifying more than one grouping variable in a list of variables handed to the argument `by=`.  For instance, here's the mean height and weight from the nhanes data, but broken down additionally by sex.

```{r}
meanWeight = groupBy(data=nhanes, 
                     by=list(H=hgroups,sex=sex),  
                     weight=mean(wgt), height=mean(hgt))
head(meanWeight)
```
Be careful not to use the same names in the by-list as in the quantities that you want to calculate for each group.

### General Observations

Notice that the operation is being identified by an assignment expression in much the same way as `transform()` uses an assigned name to create a new variable in a data frame.  The name to the left of the `=` sign gives the name of the new variable to be created.  You can name this output as you will. 

The operation specified will be performed on each group that's defined by the levels of `by`. Almost always, you will want `by` to be a categorical variable.  If you use a quantitative variable, each different value of the quantity will become it's own group, so 3.14 will be different from 3.141.  

Remember that the variables used in `groupBy` have to exist in the data frame being grouped.  

Other operations commonly performed:
* median
* spread, as with `sd()` or `IQR()`
* fraction in a range, e.g. using `prop()`
* meeting some yes or no criterion

The transformation must **always** produce a single number as an output.  If you want to compute two quantities, have two arguments describing operations.

### Table-to-Table Operations

You've now seen three operations that take a table (in R, a data frame) as an input and return a table as an output:
* `subset()` for pulling out a subset of cases
* `transform()` for creating new variables from existing ones
* `groupBy()` for aggregating cases into groups and calculating quantities on those groups.

Next week, we'll encounter a fourth operation:
* `merge()` which combines two tables to create a new table.

Keep in mind that different operations take different *kinds* of inputs and return different *kinds* of outputs:
* Some operations take a variable or variables and return another variable.  Examples: arithmetic operations such as addition, division, multiplication, and so on.
* Some operations take a variable or variables and return a number. Examples: means, sums.
* Some operations take a data frame and return a data frame. Examples: `subset`, `transform`, `groupBy`.

You should always know which you are doing.

