Taking a Defined Subset of Cases
================


```{r include=FALSE}
options(na.rm=TRUE)
require(DCF)
```

It often happens that you will be interested in only some of the cases from a data frame.  Some examples:

* You believe that some of the data is erroneous and want to exclude those cases.
* You're interested in only some of the groups represented in the dataframe, e.g. the southern states, or the elderly.

In terms of the rectangular arrangement of data in data frames, taking a subset of cases means making a selection of rows.  The variables remain unaltered, so the result of taking a subset is another data frame that has the same columns has the original.

The basic operator for taking subsets is, sensibly, `subset()`.  It takes as arguments a data frame and a criterion for making the selection.

For instance, consider the NHANES-III data on body shape.  
```{r} 
data(nhanes)
nrow(nhanes)
```
```{r results="hide"}
names(nhanes)
```
```{r echo=FALSE}
head(names(nhanes))
```

Each case is a person. Both males and females are included in the data, with the `sex` variable telling who is who:
```{r}
tally(~sex, data=nhanes)
```
Unfortunately, `sex` was encoded numerically, so you'll have to look in the documentation to find the meaning of the values.  (You can do this with `help(nhanes)`.)  It turns out that 1 stands for males, 2 for females.

Suppose, for some reason, you just want to study females.  Here's how to construct a new data frame containing only the females:
```{r}
females = subset(nhanes, sex==2)
```

Some things to notice about this simple-looking statement:

1. The criterion for selection, `sex==2`, involves the "test for equality" operator, `==`, consisting of two, adjacent equal signs.  Testing for equality, `==`, with its two characters is a completely different operation than assignment, `=`, which has just one.  There is no relationship between the two.
2. The resulting object, that is, the output of `subset` is being stored in an object.  In this example, the object is called `females` because the name is descriptive and mnemonic.  But you can use any name you like.

### Testing for Equality

You're used to declarative statements in algebra, such as $latex x=2$.  In a test for equality, `x==2`, the value of `x` is compared to the value 2.  If `x` is a collection, each element of the collection is separately compared to `2`.  The returned value is either `TRUE` or `FALSE`.  For example:
```{r}
x = c(3,2,1,2,4)  # assignment
x == 2   # test for equality
```
Some of the values in `x` are equal to 2, some aren't.  That's what the result says.

In the statement `subset(nhanes, sex==2)`, each component of the `sex` variable is being compared to 2.  Only those cases for which the result is `TRUE` are included in the subset being returned.

You can also test for equality between character strings.  For instance, in the `OrdwayBirds` data set, the `Sex` variable is encoded as characters:
```{r}
data(OrdwayBirds)
tally( ~ Sex, data=OrdwayBirds)
```
Such character labels are helpful because they remind you at a glance what their meaning is.  ("U" stands for unknown. Evidently, it's not always easy to tell what the sex of a bird is.)

To construct a subset of female birds, the statement would look like this:
```{r}
birdsf = subset(OrdwayBirds, Sex=="F")
```


### Levels of Categorical Variables

`levels` command

`%in%` for broader membership.

Negation.


Put in `drop=TRUE`.  Explain how this is a single equal sign and is assigning the value TRUE to the input `drop`.  

### Other Numerical Comparisons

Less than, greater than, 

Intervals

## When NOT to take subsets

When what you want can be done by one of the grouping operators.  These allow you to look at multiple groups at once.
```{r}
mean(wgt ~ sex, data=nhanes)
median(hgt ~ round(bmi), data=nhanes)
groupBy(data=nhanes,by=sex, roundness=mean( wst/hgt ))
```