Wake County Voters
=======================

*Instructions*

*This template includes several introductory commands and examples that you can leave as is.  Fill in your work under the "Tasks" below.*

*Do remember to put your name on the document!*


```{r include=FALSE}
source("http://www.mosaic-web.org/go/Repository/DannyKaplan/DCF_setup.R")
```      

[Wake County, North Carolina](http://en.wikipedia.org/wiki/Wake_County,_North_Carolina), has a population of over 900,000 people.  As all US counties, Wake County maintains a voter registration list.  This list is a public document. It includes over 500,000 registered voters.

For simplicity, we're going to work with a small version with just 10,000 voters.  But you can refer to the full data if you want.

To read in the sample of size 10,000:
```{r}
data(WakeVotersSmall)
```

### Simple Group Operations

You can use `groupBy` to do simple count calculations.  Some examples:
#### How many voters of each party?
```{r}
groupBy(data=WakeVotersSmall, by=party, count=length(party))
```

#### How many voters of each race?

```{r}
groupBy(data=WakeVotersSmall, by=race, count=length(party))
```

Notice that there are a handful of people whose race is listed as blank.  

#### How many voters of each sex and race?

When there are two (or more) factors involved, the `by=` argument should be set to be a list, like this.
```{r}
both = groupBy(data=WakeVotersSmall, by=list(race,gender), count=length(party))
head(both)
```

#### A bit of data cleaning

There are so few `LIB` in the data that it's inconvenient to include them for general purposes.  Similarly, with the people whose race is listed as blank.  To simplify your later work, remove all the `LIB` and blank race people from the data.  We'll call the resulting, somewhat simplified data `voters`.
```{r}
# do a bit of cleaning
voters = droplevels(subset(WakeVotersSmall, party != "LIB" & race != ""))
```

### Creating Age Groups

Age is sensibly listed in terms of years.  Sometimes it's convenient to divide a quantitative variable into groups.  Let's add a new variable, `AgeGroup` that splits up the age.
```{r}
voters = transform(voters, 
                   AgeGroup=cut(Age, breaks=c(0,30,60,100) ) )
```

### Demographics of Party Affiliation

It's widely believed that different sex, age and racial groups have different political preferences.  

You can use `groupBy` to count how many in each party, broken down by one or more variables.  For instance:
```{r}
byAge = groupBy(data=voters,
                 by=list(party,AgeGroup))

byRace = groupBy(data=voters,
                 by=list(party,race))
```

Here are some graphs showing the relationship between party affiliation and age:
```{r echo=FALSE}
ggplot(data=byAge,aes(x=AgeGroup,y=count,fill=party))+
  geom_bar(stat='identity',position=position_fill(width=.9)) + 
  theme(axis.text.x=element_text(angle=60,hjust=1)) 
ggplot(data=byAge,aes(x=AgeGroup,y=count ,fill=party))+
  geom_bar(stat='identity',position=position_dodge(width=.9)) + 
  theme(axis.text.x=element_text(angle=60,hjust=1))
```

### Task 1

Use `mBar(byAge)`  to make the graphs above.  You'll have to figure what variables go in what positions.

*Since the commands are given above, there's little point in your making the graphs again.  But explain which variable takes which role in the graphs.*

### Task 2

What's the relationship between party affiliation and race?  Does it differ by the sexes?

Make one or more appropriate graphs and provide a narrative giving your conclusion.

```{r}
# put your graphics commands here
```

*Put your narrative interpretation of the graphs here.*

### Task 3

Does the distribution of races differ by age?  

```{r}
# put your graphics commands here
```

*Put your narrative interpretation of the graphs here.*

### Task 4

Does the distribution of name suffixes (`name_sufx`) differ by sex?  By race?

```{r}
# put your graphics commands here
```

*Put your narrative interpretation of the graphs here.*

------------------

Compiled on `r I(date())`