Regular Expressions
========================================================

```{r include=FALSE}
require(DCF)
#startTemplate()
```

We have seen how we can search for data rows when a string value
for a variable mathes exactly, like this:

```{r eval=FALSE}
sparrow1 <- subset(OrdwayBirdsOrig, SpeciesName == "Field Sparrow")
```

Now we will explore how we can go beyond exact string matching, using some narly, yet powerful constructs called regular expressions, which can depict a general pattern of text that we wish to compare to a string.

First, let's go through [a tutorial on creating regular expressions](http://regexone.com/)

Here is how we could get a subset of just the names of the Sparrows, using
the grepl operator, which works like ==, in that it returns TRUE when the pattern given matches the data value for the varible called SpeciesName:

```{r eval=FALSE}
sparrowNames <- subset(OrdwayBirdsOrig, grepl(".*Sparrow$", SpeciesName), select="SpeciesName")
```

How would you change it to filter out only Song and Chirping Sparrows?

Recall that we can get all the data variables by elimiating the select, like this:

```{r eval=FALSE}
sparrowBirds <- subset(OrdwayBirdsOrig, grepl(".*Sparrow$", SpeciesName))
```

The Release column variable has data that looks like this:

to no. meadow
to n. meadow edge
to north meadow

How many different variations of 'north meadow' (or something like it) are there for the Release data?

Can we now find a way to now get Sparrows who flew to the 'north meadow' but not to the 'north meadow edge'?


