Regular Expressions
========================================================

```{r include=FALSE}
require(DCF)
#startTemplate()
```

We have seen how we can search for data rows when a string value
for a variable mathes exactly, like this:

```{r}
data(OrdwayBirdsOrig)
sparrow1 <- subset(OrdwayBirdsOrig, SpeciesName == "Field Sparrow")
```

Now we will explore how we can go beyond exact string matching, using some narly, yet powerful constructs called regular expressions, which can depict a general pattern of text that we wish to compare to a string.

First, let's go through [a tutorial on creating regular expressions](http://regexone.com/)

Here is how we could get a subset of just the names of the Sparrows, using
the grepl operator, which works like ==, in that it returns TRUE when the pattern given matches the data value for the varible called SpeciesName:

```{r}
sparrowNames <- subset(OrdwayBirdsOrig, grepl(".*Sparrow$", SpeciesName), select="SpeciesName")
```

How would you change it to filter out only Song and Chirping Sparrows?

Recall that we can get all the data variables by elimiating the select, like this:

```{r}
sparrowBirds <- subset(OrdwayBirdsOrig, grepl(".*Sparrow$", SpeciesName))
```

The Release column variable has data that looks like this:

to no. meadow
to n. meadow edge
to north meadow

How many different variations of 'north meadow' (or something like it) are there for the Release data?

Can we now find a way to now get Sparrows who flew to the 'north meadow' but not to the 'north meadow edge'?

Transforming Data
-----------------

There is another useful wat to use regular expressions in R, using operators available in the package called stringr.  As an example, we can extract a portion of a string that matches a pattern by using the operator called str_extract, as in this example:

```{r}
require(stringr)
sparrowTypes = transform(sparrowBirds, 
                         SparrowType = str_extract(SpeciesName, "^\\w+"))
# Get a subset of variables
sparrowTypes = subset(sparrowTypes, 
                      select=c(SpeciesName, SparrowType, Age, BandNumber))
```

Note that we are first creating a new column in the data whose name is SparrowType and that the portion of the pattern that matches first in SpeciciesName is saved as a new data element.  

Try this:

Can we use this method to get the 2-3 letter tags used at the beginning of the 'Age' data and place them in a new column? 

This is a common operation: extract data that has been combined in one column and place it in spearate columns.  However, this operator has its limitations, because it finds the first occurance of a pattern and extracts that.  Suppose we want to get the description within the parentheses in Age.  This will get it with the parentheses:

```{r}
sparrowTypes = transform(sparrowTypes, AgeDescription = str_extract(Age, "\\(.*\\)"))
```

It is a bit trickier to remove the parentheses.
