Foreign Direct Investment and the Gender Wage Gap
========================================================

```{r include=FALSE}
require(DCF)
require(plyr)
```

This is from a project done by Macalester student Pukitta Chunsuttiwat '15 for a class in Economic Restructuring in Latin America, taught by Prof. Raymond Robertson.

The issue is whether the amount of foreign direct investment in Latin American countries has an effect on the gender wage gap, specifically, the hypothesis that higher FDI leads to a smaller gap.

Variables:
* FDI --- Foreign Direct Investment (USD)
* Wage gap using data from the UN
* Controlling for percentage of female employment

The data come from the World Bank, several files created by Pukitta from the [World Bank site]().  These data files have been downloaded to this directory.

```{r}
setwd("CaseStudies/ForeignInvestmentAndWageGap")
```

### Countries of interest

```{r}
CofI = c("Brazil", "Costa Rica", "El Salvador", "Mexico", "Panama", "Paraguay")
### Read in the UN data on the Wage Gap

```{r}
gap = fetchData("UN-Wages-flattable.csv")
gap = subset(gap, Country.or.Area %in% CofI & Coverage=="Employees")
gap = subset(gap, SourceID=="# 0" | (Country.or.Area=="Mexico" & SourceID=="# 1"))
vars = c("Country.or.Area", "Year", "Value")
women = subset(gap, subset=Sex=="Women", select=vars)
men = subset(gap, subset=Sex=="Men", select=vars)
women= rename(women, c("Value"="WageF"))
men = rename(men, c("Value"="WageM"))
gapboth = merge(women,men)
gapboth = transform(gapboth, WageRatio=WageF/WageM)
```

We're interested in the relative wages of men and women:
```{r}
xyplot( WageRatio ~ Year | Country.or.Area, data=gapboth)
```

### FDI by country and year
The wages are in local currencies.  We want to put them on a common footing.  For this, we'll need a conversion for each country for each year

```{r}
FDI = fetchData("FDI World Bank.csv")
foo = melt(FDI, id.vars=c("Country.Name"), 
           measure.vars=paste("X",1999:2008,sep=""))
foo = rename(foo, c("variable"="Year", "value"="FDI"))
FDI = transform(foo, Year=as.numeric(as.character(gsub("X","",Year))))
```

```{r}
xyplot( log10(FDI) ~ Year | Country.Name, data=FDI)
```

### Employment to Population Ratio for Females

```{r}
ratio = fetchData("Employment to Pop Ratio Female.csv")
ratio = subset(ratio, Indicator.Name=="Employment to population ratio, 15+, female (%)")
ratio = subset(ratio, Country.Name %in% CofI)
foo = melt(ratio, id.vars=c("Country.Name"), 
           measure.vars=paste("X",1990:2012,sep=""))
foo = rename(foo, c("variable"="Year", "value"="EmploymentRatio"))
ratio = transform(foo, Year=as.numeric(as.character(gsub("X","",Year))))
```

```{r}
xyplot( EmploymentRatio ~ Year | Country.Name, data=ratio)
```


### Put the tables together

```{r}
gapboth = rename(gapboth, c("Country.or.Area"="Country.Name"))
final = merge(FDI, gapboth)
final = merge(final,ratio)
```

### Construct the model:

```{r}
mod = lm( WageRatio ~ FDI + EmploymentRatio + Country.Name + Year, data=final)
anova(mod)
```