Exercises: Week 5
========================================================

```{r include=FALSE}
require(DCF)
startTemplate()
```

#### Data and Computing Fundamentals


### Gapminder Data

For your reference, here are commands to read in a set of Gapminder files, taking the mean value of the variable for year 2000 and later.
```{r message=FALSE}
varNames = c("Alcohol", "Fertility", "TrafficDeathRate")
res = NULL  # start with no data
for (XXX in varNames) {
    gapdat = fetchGapminder(paste("Gapminder/", XXX, ".csv", sep = ""))
    gapdat = subset(gapdat, Year >= 2000)
    command = parse(text = paste("mean(", XXX, ")"))
    oneVar = groupBy(gapdat, by = Country, value = eval(command))
    oneVar = rename(oneVar, c(value = XXX))
    # Outer join with existing data
    if (is.null(res)) {
        res = oneVar
    } else {
        res = join(res, oneVar, by = "Country", type = "full")
    }
}

```

#### Exercise 1

Make a parallel coordinates plot involving `Fertility`, `Alcohol` and `TrafficDeathRate`.  Show the plot and interpret it as regards the connection between per capita alcohol consumption and traffic death rate.

<!-- Notes for a potential exercise.  What's the relationship between alcohol consumption and traffic deaths.  It appears that fertility has a stronger connection --- the higher the per capita alcohol consumption, the lower the traffic death rate.  Ecological paradox: that this pattern holds true for countries doesn't mean that it will hold true for individual people. -->

```{r eval=FALSE}
res = transform(res, TD=log(TrafficDeathRate))
parallelPlot( ~ Fertility - Alcohol + TD, data=res,axes=TRUE)
```

#### Exercise 2

Pull out a set of Gapminder variables that you think are important 
in showing health risks.  
* Use them to cluster countries.  
* Make a map showing the geographical distribution of the health risk clusters.  
* Use scatter plots or parallel coordinate plots to give a meaningful, descriptive name to each cluster.


Do the same for variables that indicate

1. Pro-feminist policies
2. Economic advancement.
