

```{r include=FALSE}
require(DCF)
startTemplate()
```

### installing the mclust

Make it a required part of DCF?  

```{r eval=FALSE}
faithfulMclust <- Mclust(faithful)
plot(faithfulMclust,what="classification")
classError
hcTree() and cl() # to cluster into different numbers of groups and return the index of the classification.
```

### A brush-able, interactive parallel coordinates plot: [here](http://syntagmatic.github.com/parallel-coordinates/)


An example of a parallel plot: characteristics of cars:
```{r}
parallelPlot(mtcars, ~ wt +  cyl + disp + hp - mpg,lwd=5,axes=TRUE,alpha=.5)
```
Which car is the sports-car?

Which are the econo-boxes?

What exceptions are there to the rule: lighter weight means better gas mileage?

### Exercise on parallel coordinate plots

Use the kids data from the `fastR` package.  Use popularity as the first variable so that it controls the coloring.  At a glance, what is associated with popularity in the individual variables.  Can you find a set of variables and an inversion that tends to keep lines parallel?

Maybe some of the other multi-variable things from `fastR`, e.g. `concreteAll`

### Outline



#### II. Density and clustering

A. How to show large numbers of points in a scatter plot, e.g. `hexbin`, low `alpha` (which I'll have them do programmatically), faceting (also programmatically)
B. Unsupervised clustering: bird wing widths 
    (next week, supervised clustering and tree models, cross validation)
C. Clustering in two dimensions, where you can see what's going on
D. Clustering in three and more dimensions.
E. Bar plot of the bird cluster groups by species --- how are we doing?

#### III. Parallel coordinates for hypothesis formation

#### IV. Many variables.

##### A. Programming techniques 

Write `Index.csv` file for the Gapminder directory.  List the file names.

Have them write a function that reads in a data file, subsets out the years from 2000-20005, and takes the average from 2000-2005, 

Join ("outer") this with another variable.

Programming: loops
1. An accumulator: `res=NULL`
2. The loop: over the files you want to read
3. Accumulation, join the new file to the old one
    Technique: assuming that `join` doesn't do the right thing with NULL, use an `if/then` structure
    
Assignment: do parallel coordinates and unsupervised clustering on the country data to assign a group to each one. Then make a map of the groups.  Interpret the results using your geopolitical knowledge.


Exercise: More advanced: Cut year into decades. Group by country,decade and take the mean or max of the values.  Later on, take subsets to get individual decades

##### V. New Data --- nutrition by country