Computational Summary
====================

An evolving document showing the set of commands and command types for students in the DCF course.

```{r include=FALSE}
require(ggplot2)
require(gcookbook)
require(plyr)
require(mosaic)
require(rworldmap)
```


### Graphics

#### Day 1

```{r}
library(gcookbook)
p <- ggplot(heightweight, aes(x=ageYear, y=heightIn, color=weightLb)) + geom_point(size=3)
library(scales)
p + scale_colour_gradientn(colours=c("darkred","orange", "yellow", "white"))

#### Day 5 Density

```{r}
p <- ggplot(faithful, aes(x=eruptions, y=waiting))
p + stat_density2d() + geom_point()
```

#### Day 6 Classification

```{r}
library(gcookbook)
c2 <- subset(countries, Year==2009)
c2 <- c2[complete.cases(c2),]
set.seed(201)
c2 <- c2[sample(1:nrow(c2),25),]
rownames(c2) <- c2$Name
c3 <- scale(c2[,4:7])
hc <- hclust(dist(c3))
plot(hc, hang=-1,main="",xlab="",ylab="")
```
### Day 1

Scatter plots from the UN FAO population and country size data
```{r}
FAO <- read.csv("FAO_DATA/FAO-simple.csv")
ggplot(data=FAO)+geom_point(aes(x=Arable.land/Country.area,y=I(Rural.population/Urban.population)))+aes(colour=log10(Country.area))+aes(size=log10(Total.Population...Both.sexes))+scale_x_log10()+scale_y_log10()
```


### Day 2

Bird data, voter data
```{r}
birds <- read.csv("Data/Ordway/christman-short.csv")
count <- data.frame(tally( ~species.name, data=birds))
names(count) <- c("species.name","species.count")
hoo <- join(birds, count, by="species.name")
hoo <- droplevels(subset(hoo, species.count > 200))
counts <- data.frame(tally( ~species.name | Month, format="count",margins=FALSE, data=hoo))
ggplot(data=counts,aes(x=species.name,y=Freq ,fill=Month))+geom_bar(stat='identity',position=position_fill(width=.9)) + theme(axis.text.x=element_text(angle=60,hjust=1)) 
ggplot(data=counts,aes(x=Month,y=Freq ,fill=species.name))+geom_bar(stat='identity',position=position_stack(width=.9)) + theme(axis.text.x=element_text(angle=60,hjust=1)) 
```

Questions:
1. Which species are migratory?
2. Which species winter over? Which species summer over?
3. Are the sexes about equally represented?  Does sex ratio depend on species?

Do a bit of research on each of the species?  According to Wikipedia --- what questions are worth asking: migratory?  insect eating?  

Is sex related to age?

Do the species vary by time of day?

Pick a month of the year.

### Day 3

Maps
* Use `rworldmap` package for world maps
* Perhaps `maps` or `ggplot2` for US maps

```{r}
FAO$fracUrban <- with(FAO, Urban.population/Total.Population...Both.sexes)
sPDF <-joinCountryData2Map( FAO, joinCode="NAME",nameJoinColumn="Country")
mapCountryData(sPDF, nameColumnToPlot="fracUrban")
```

```{r}
library(maps)
us <- map_data("state")
world <- map_data("world")
crimes <- data.frame(state=tolower(rownames(USArrests)),USArrests)
crime_map <- merge(us, crimes, by.x="region", by.y="state")
crime_map <- arrange(crime_map, group, order)
ggplot( crime_map, aes(x=long, y=lat, group=group, fill=Assault)) + geom_polygon(color="black") + coord_map("polyconic")
```