```{r include=FALSE}
require(DCF) # leave this chunk alone: boilerplate
```

United Nations Food & Agriculture Organization: Week 1
=======================

### Your Name Goes Here


The data on area, land use, and population are from the UN FAO records for 2009.
```{r}
data(FAOsimple) # Read in the data: boilerplate
```

#### Editing this document

The point of this template is to give you a working document that you can modify in steps to create your final report.  Of course you will want to delete the template contents and substitute your own.

R chunks labelled "boilerplate" should be left alone.


### Techniques
#### Creating a New Variable

You can make new variables by combining existing variables.  Here's the fraction of the population engaged in agricultural work, a simple fraction:
```{r}
FAOsimple = transform(FAOsimple,
  popAgFrac=Agricultural.population/Total.Population...Both.sexes)
```

#### Making a Simple Scatter Plot

Here's an example of a plot showing the UN FAO data:
```{r fig.width=8,fig.height=4,warning=FALSE}
ggplot(data=FAOsimple)+
  geom_point(aes(x=Land.area,y=Agricultural.area),alpha=.5)+
  aes(color=popAgFrac,
      size=log10(Total.Population...Both.sexes))+
  scale_x_log10()+scale_y_log10() 
```

This graph of each country's agricultural area versus total land area shows some aspects of the relationship between land use and population.
* There is a sharp diagonal boundary to the cloud of points because a country's agricultural area cannot be larger than its land area.  
* Countries with an agricultural area much smaller than the land area tend to have only a small fraction of the population in agriculture.  
* Countries with agricultural area about the same as the land area (near the diagonal) tend might have either a small or a large fraction of the population in agriculture.


Once you have a basic plot, you can easily substitute names of new variables for old, without having to go through the user interface.


----------------------
Compiled on `r I(date())`
