```{r include=FALSE}
require(DCF)
options(na.rm=TRUE)
```

### Data and Computing Fundamentals

Data on body shape and diabetes are contained in the `nhanes` dataframe.

```{r eval=FALSE}
data(nhanes)
```

For a description of the variables, see
```{r eval=FALSE}
help(nhanes)
```

#### QUESTION: 

Describe the relationship between height and weight.  Is there reason to think that it's different for the two sexes?

#### QUESTION

Describe the relationship between weight and BMI.  Is it different from the two sexes?  Where do the people with diabetes show up?

#### QUESTION

Is there a relationship between BMI and age?  Where do the people with diabetes show up?

#### QUESTION

Is cholesterol level a good predictor of the development of diabetes?

#### TASK: 

Calculate body-mass index according to the formula $\frac{m}{h^2}$ and see how it corresponds to the body mass index in the data.  (Hint: You can plot out one versus the other.)

#### TASK:

Come up with a hypothesis of your own and address it with a graphic.
