### Data and Computing Fundamentals

Week 1: In-Class Data Exploration
================

```{r include=FALSE}
require(DCF)
options(rm.na=TRUE)
```

Read the the simplified version of the United Nations Food and Agriculture Organization dataset using
```{r}
data(FAOsimple)
```
Look at the names of the variables and make sure you understand what is the meaning of each of them.

You're going to be making scatter plots using mScatter(). To generate the graphics in this document, remember to cut-and-paste the command output of mScatter() into a fenced R command in this document, e.g.

<pre>
ggplot(data=FAOsimple)+geom_point(aes(x=Total.Population...Both.sexes,y=Agricultural.area))
</pre>

#### QUESTION:

Is there a clear relationship between area of a country and the total population?

#### QUESTION:

Is there a clear relationship between the fraction of a country's total land that is arable and the fraction of a country's total population that is engaged in agricultural work?

#### TASK:

Construct new variables
* Calculate population density
    * population per area
    * population per agricultural area
* Add up the male and female population and verify that it matches total population.  (You can make a new variable and plot it against the original total population.)


#### TASK:

Pick out some variables that you think ought to be related.  Describe the relationship you hypothesize and then demonstrate whether or not your hypothesis is consistent with the data.

#### TASK:

Show a situation where the use of logarithmic axes helps to illustrate a relationship.

