```{r include=FALSE}
require(DCF)
options(na.rm=TRUE)
```

### Data and Computing Fundamentals

Week 1: Out-of-Class Activity
-----------------


## Data on the Web

Look at the various data sets available at <http://www.gapminder.org/data/>.

1. Based on the names of the data sets available, construct several hypotheses that you would like to test using this sort of country-by-country data.  Later in the course, you'll learn how to assemble such data in a form you can use for your analysis.
2. Look at one of the data sets (by pressing the little "View" magnifying-glass icon).  What are the cases?  What are the variables?

## Body Shape Data

Read the the simplified version of the NHANES dataset using 
```{r}
data(nhanes)
```

You can use `help(nhanes)` to see t
Look at the names of the variables and make sure you understand what is the meaning of each of them.

You're going to be making scatter plots using `mScatter()`.  To generate the graphics in this document, remember to cut-and-paste the command output of `mScatter()` into a fenced R command in this document.

#### Create a Sample

It takes several seconds to generate a graph using this number of data points.  To speed things up, take a random sample of 2000 people and develop your graphs with that.  
```{r}
small = sample(nhanes,2000)
```

Then, when you know exactly what you want, you can translate your commands to use the whole data set, if appropriate.

------------

#### QUESTION: 

Describe the relationship between height and weight.  Is there reason to think that it's different for the two sexes?

#### QUESTION

Describe the relationship between weight and BMI.  Is it different from the two sexes?  Where do the people with diabetes show up?

#### QUESTION

Is there a relationship between BMI and age?  Where do the people with diabetes show up?

#### QUESTION

Is cholesterol level a good predictor of the development of diabetes?

#### QUESTION

Come up with a hypothesis of your own and address it with a graphic.

