Week 3: Exploring Gapminder Data through Maps
===================

```{r include=FALSE}
require(DCF)
startTemplate()
```

### Data and Computing Fundamentals



#### Exercise 1

Make a world map in which the population of each country is indicated by a color.

#### Exercise 2

Make a world map in which the population of each country is indicated by the size of a circle.

#### Exercise 3

Make a world map in which the population density of each country is indicated by a color.

#### Exercise 4

Make a map, like this, that correlates traffic deaths with fraction of roads that are paved.  (See `PavedRoads.csv` in the Gapminder data.)

```{r echo=FALSE}
traf = fetchGapminder("Gapminder/TrafficDeathRate.csv")
tmean = groupBy(traf, by=Country, deaths=mean(TrafficDeathRate))
roads = fetchGapminder("Gapminder/PavedRoads.csv")
rmean = groupBy(roads, by=Country, paved=mean(PavedRoads))
both = join( rmean, tmean )
both = transform(both, deathQ=ntiles(deaths,5))
{.s. <- joinCountryData2Map(both, joinCode='NAME',nameJoinColumn='Country' ); par(mai=c(0,0,0.2,0), xaxs='i',yaxs='i') ;.mp. <- mapBubbles(.s., nameZColour='deathQ',nameZSize='paved',mapRegion='eurasia')} 
```

* What sort of correlation is there between fraction of roads paved and traffic death rates?  
* Are there countries that stand out of the trend?  Which ones

Make a scatter plot of the traffic death rate, fraction of paved roads, and alcohol consumption per capita.  Compare the map portrayal to the scatter plot --- which one makes it easier to see various patterns.
