Week 3 Notes
===============

```{r include=FALSE}
require(DCF)
startTemplate()
require(xtable)
```

Who's cool in the [*Wall Street Journal*](http://online.wsj.com/article/SB10001424127887323478304578332850293360468.html?mod=WSJ_hps_RightRailColumns#)

## Todays' Topics

* Accessing Gapminder data on the web.
* Wide versus narrow data
* Making maps

The maps we make won't be as sophisticated as this one from the New York times, but are still pretty impressive. [NYTimes Map of Hispanic population in California](http://www.nytimes.com/interactive/2013/02/20/us/hispanics-californias-next-majority.html?hp)

## The Gapminder Data site

They offer analysis software, as do many data-oriented sites.  But it's much better to learn to use a general-purpose and customizable tool like R.  So we're going to use their data.

* Pick a data set.
* Look at the data file and the documentation.
* Download a CSV file to your computer
* Get the link to the documentation
* Format an entry in the table form and email your file to me.
<code>  
    Short description | \[link](documentation link) | \`Gapminder/Filename.csv\`
</code>

## Wide versus Narrow

Read in one of the data files, e.g. `dat = fetchData()` and navigate.  Or...
```{r}
dat = fetchData("Gapminder/Fertility.csv")
```

### Simple operations on the data

How would you do these things?
* How many countries?
* How many years?

But could you do these things?
* How many countries for each year?
* How many years for each country?

### The Narrow Format

Here's a special function that reads things in in narrow format:
```{r}
fertility = fetchGapminder("Gapminder/Fertility.csv")
```

Do each of these things:
* How many countries for each year?  
* How many years for each country?
From the answers to these, you can figure out the answers to these:
* How many countries?
* How many years?


## Getting one value per country

Use `groupBy()` with `by=Country` and pull out some summary of the values, e.g., `min`, `mean`, `median`, `sd`

* What's a *standard deviation*?  Plot values on a number line.  The mean is the center of mass.  The median is the point such that half are to the left and half to the right.  The standard deviation is the typical distrance from the mean.


## Draw a Map of the Data

Reduce your data down to one value per country.

Use `mWorldMap()`

## Discrete colors

## Quantitative to Categorical

`ifthen`, `cut`, `ntiles`

* How even-numbered are the groups?
* Widely divergent data: ntiles does something like this automatically.

## Joining two data files

* Very simple --- just use `join`

## Maps and Scatter Plots

* Use `mWorldMap` to generate the command.

## Animating Graphics: The Basics

How has fertility changed over the years:
```{r}
fert = fetchGapminder("Gapminder/Fertility.csv")
co2 = fetchGapminder("Gapminder/CO2emissions.csv")
both = join(fert,co2)
```

Now plot out one versus the other for a single year:
```{r eval=FALSE}
manipulate({thisyear = subset(both, Year==year)
ggplot(data=thisyear)+geom_point(aes(x=CO2emissions,y=Fertility))+scale_x_log10(limits=c(1,1e9))+ylim(0,10)},year=slider(min=1950,max=2010,label="Year")) 
```

Things to add:
* Country population as a size
* region as a color

