Wide versus Narrow Data Formats
======================

```{r include=FALSE}
require(DCF)
startTemplate()
require(xtable)
```

The [Gapminder](www.gapminder.org) data are stored as individual files for each variable.  The format is **wide** --- each year is its own column.

To read in a data file, use `fetchData()` in the ordinary way.  For instance, per capita alcohol consumption each year (in litres per adult, where an adult is defined as a person aged 15 years and over):
```{r}
alcWide = fetchData("Gapminder/Alcohol.csv")
```

Of course, you're tempted to look immediately at the data file, but it's fairly large:
```{r}
nrow(alcWide)
names(alcWide)
```
There are `r nrow(alcWide)` countries.  There are data for `r length(names(alcWide))-1` years.  Altogther, that's `r nrow(alcWide)*(length(names(alcWide))-1)` entries.

The variable names are a bit strange, but you can see that there is a different variable for each year, even though the value is the same thing: alcohol consumption in litres.  

Here's a small part of the data table: a few countries over a few years.
```{r echo=FALSE,results="asis"}
small <- alcWide[156:165,c(1,14:25)]
names(small) <- c("Country","1997", "1998", "1999", "2000", "2001", "2002", "2003", "2004","2005","2006","2007","2008")
print(xtable(small,caption="Alcohol Consumption"),type="html")
```

In this particular data set, each case is one country, each variable is alcohol consumption for each year.

This is what's called a **wide** format for data.  A wide format is intended to make it easy to pull out all the data for one case as a `subset` operation, or all the data for one variable (in these data, one year) as a `projection` operation.  For instance:
```{r}
subset(alc, X=="Syria")
```

The most notable feature of this particular wide format is that most of the values are missing.  Presumably, that's because no reliable estimate of alcohol consumption was available for most countries in most years.

Another possible format for these data is called **narrow**.  The narrow format has a different idea of what's a case.  Rather than treating each country as a case, the case is each country in a particular year.  This is what a narrow format looks like:


```{r echo=FALSE, results="asis",echo=1}
alc = fetchGapminder("Gapminder/Alcohol.csv", value.name="Alcohol")
print(xtable(alc[1:10,]),type="html")
```

All the alcohol measurements are now in one variable, the years in another.  You can still easily grab the data for one year or one country, e.g.
```{r results="hide"}
subset(alc, Country=="Russia")
subset(alc, Year==1990)
```

The `groupBy()` function works well on narrow data and allows you to consolidate data across countries or years.

Here's a fairly elaborate example using operations that may be unfamiliar: Finding the country with the maximum alcohol consumption in each year:
```{r}
groupBy(alc, by=Year, 
        Country=as.character(Country[which.min(Alcohol)]), 
        Consumption=min(Alcohol))
```

Why do you think that Sweden and Russian are often found to be the countries with the **minimum** alcohol consumption?  Hint: It's not because people in all other countries drank more in those years.

### Comparing multiple variables

The narrow format is particularly useful when you want put two or more variables side by side. Suppose you read in two variables:
```{r echo=FALSE, results="asis",echo=1:2}
alc = fetchGapminder("Gapminder/Alcohol.csv", value.name="Alcohol")
liver = fetchGapminder("Gapminder/LiverCancerIncidenceMale.csv",value.name="LiverCancerM")
print(xtable(alc[1:5,]),type="html")
print(xtable(liver[1:5,]),type="html")
```

Suppose you want to do a quick exploration of the potential link between liver cancer and alcohol consumption.  A scatter plot of the incidence of liver cancer against alcohol consumption might be informative.  To make this, you have to put the two variables side by side.  Of course, you'll need to match up the corresponding cases.  This is not just a matter of matching the first row in one data frame to the first row in the other; you don't want to put alcohol in Russia in 1985 against liver cancer in Denmark in 1953.  Instead, you want to join the tables in a sensible way, matching up the corresponding cases.

Doing such matching requires that there be identifiers that mark the individual cases.  In for these data, the unique identifier is country and year --- for each variable there is only one entry for each country and year.  Later on, when you study the **join** operation, you'll see that there are choices to be made.  But for data like these, organized in very similar ways, the default operation is pretty simple.

```{r echo=1,results="asis"}
both = join(alc,liver)
print( xtable( both[14:19,]),type="html")
```

Notice that when there is a case in one table but no corresponding case in the other --- for instance, Ukraine in 1993 --- the variable is listed as `NA`.


