Medicare Spending 
========================================================

```{r include=FALSE}
require(mosaic)
# require(DCF)
```

In May 2013, the [Centers for Medicare and Medicaid Services](http://www.cms.gov/Research-Statistics-Data-and-Systems/Statistics-Trends-and-Reports/Medicare-Provider-Charge-Data/index.html) released data on the price of health services at hospitals around the US.  This attracted press attention.  For example, a headline in the New York Times read ["Hospital Billing Varies Wildly, Government Data Shows"](http://www.nytimes.com/2013/05/08/business/hospital-billing-varies-wildly-us-data-shows.html).  A [map with the article](http://www.nytimes.com/interactive/2013/05/08/business/how-much-hospitals-charge.html?_r=0) shows the geographic disparities in hospital charges.  

* Read the New York *Times* article to familiarize yourself with the issue.

The data themselves are available as a [zip-compressed CSV file](http://www.cms.gov/Research-Statistics-Data-and-Systems/Statistics-Trends-and-Reports/Medicare-Provider-Charge-Data/Downloads/IPPS_DRG_CSV.zip).  

DRAFTY: This file format happens to be wasteful in storage and speed for reasons we will discover later in the semester.  You can read in the data frame, called `med` with: (PUT THIS IN THE PACKAGE or on a fetch repository)
```{r}
load("CMS-data-2013-05-06.rda)
```

### Some explanations 

From the [FAQ for the site](https://questions.cms.gov/faq.php?id=5005&rtopic=2038
)
* What is the difference between “Average Charges” and “Average Total Payments”?
    “Average Charges” refers to the what the provider bills to Medicare. “Average Total Payments” refers to what Medicare actually pays to the provider.

* `Total.Discharges` is the number of cases.

### To Do

* Get population by city
* Get population by state
* Get median income by city

### Topics

#### Simple Summary

1. How many Direct Recovery Groups are there?
2. How many hospitals are there?
3. How many cities are there?

#### Ordering

1. Which are the most commonly performed DRGs in these data?
2. Which are the highest price DRGs in these data?

#### Maps

1. Make a map of the state-wide charge in DRG 


#### Classification

1. Divide up DRG/Hospitals into ntiles by the number of cases.  Is there a trend for the amount charged by number of cases?  Is there a trend for the total payments by the number of cases?
1. Pick out two DRGs that are common but are different from one another (e.g., knee replacement and heart failure).  Make a map of the ratio of case numbers and of costs.
2. Pick out the 10 most common DRGs nationally.  Do clustering on the proportion of all cases for each state of these DRGs.  Make a map of cluster number.

#### Grouping

1. How much money was paid to each provider?
    * O





#### Join

1. Population
    1. Add population of each state. (example)
    2. Add population of each city. (exercise)
2. Variation in each DRG (say, IQR or sd)


### Some tasks

1. How many Direct Recovery Groups are there?
2. What is the most expensive group (on average and in the median)?
3. What is the most expensive state (on average and in the median)?
4. What is the most expensive city (on average and in the median)?
5. What DRG has the largest spread of values? (use the sd)
6. As a proportion from the median, which DRG has the largest spread of values? (use the sd divided by the mean)


Later on (when we do maps): 
1. Find the z-score for each DRG.  Make a map of the average z-score by state.

When we do join:
1. Look at how the data are not normalized.  Present a normalized version of the data and show how much smaller it is.  Link in some other variable relating to zip code, for example per capita income.


