Review Activity --- Week 6 --- Answers
==================

### Data and Computing Fundamentals

```{r include=FALSE}
require(DCF)
startTemplate()
```

## Introduction


```{r}
load(url("http://www.mosaic-web.org/go/datasets/CFPB.rda"))
```
Once this is done, the CFPB database will be stored in a data frame named `CFPB`.  You can look at the various variables to get an idea of the kinds of financial services included in the data.  For instance
```{r}
answered = groupBy(CFPB, by=list(Product,Submitted.via))
```

### Warm Up

Make a bar chart that tells an interesting story about the CFPB data.
Student-loan complaints are more likely to be submitted via email:
```{r}
ggplot(data=answered,aes(x=Product,y=count ,fill=Submitted.via))+geom_bar(stat='identity',position=position_fill(width=.9)) 
```


## Main Activity

```{r}
load(url("http://www.mosaic-web.org/go/datasets/zipPop.rda"))
## Fix the zip code
zipPop = transform(zipPop, 
                 Zip=as.numeric(gsub(" zip code","",as.character(Zip))))
```



1. Find the total population represented in the `zipPop` file.  How does it compare to the standard estimates of the US population.
2. Find the population of each state based on the `zipPop` file.
3. Nationally, find out how many CFPB complaints there are per capita (using the population estimate from `zipPop`).  Answer: it's about 3.25 per 10,000 people.  You should give a statement that you used to compute this number.

```{r}
with(zipPop, sum(Population, na.rm=TRUE))
states = groupBy(zipPop, by=State, Population=sum(Population, na.rm=TRUE))
nrow(CFPB)/with(states, sum(Population))
```
### Combining the ZIP-code with the CFPB data

```{r}
CFPB = rename(CFPB, c("ZIP.code"="Zip"))
```

And the join ...
```{r}
both = join(CFPB, zipPop, by="Zip")
```

**Question**: Why was `CFPB` listed first in the `join()` command?

Alas, not all the complaints come from a ZIP code that's represented in our ZIP-code population data:
```{r}
groupBy(both, by=is.na(Population))
```

Ideally, we would hunt down the missing ZIP codes.  But, for now, let's just get rid of the missing cases.
```{r}
both = subset(both, !is.na(Population))
```

### State-by-State Summaries

Let's see how many complaints come from each state and compare that to the expected number of complaints, given the state's population.  First, the number of complaints in each state.

```{r}
countByState = groupBy(both, by=State)
```

Next, the population of each state
```{r}
popByState = groupBy(zipPop, by=State, 
                     Population=sum(Population,na.rm=TRUE))
```

**Question**: By find state-by-state population from the `zipPop` data rather than `both`?

Join the complaint count and the population:
```{r}
states = join(countByState, popByState, by="State")
```

Find the number of complaints per capita nationwide.  This will be different from the 3.25 per 10000 because we threw away some of the complaints --- the ones that don't match to an entry in our `zipPop` data.
```{r}
perCapita = nrow(both)/with(popByState, sum(Population))
perCapita
```

Use the number you found to find the expected number of complaints per state. 
```{r}
states = transform(states, expected=perCapita*Population)
```

Are there states that have much more than the expected number of complaints?  Which ones?  Are there states that have many fewer than the expected number of complaints?  Which ones?

```{r}
states = transform(states, 
                   upper=qpois(0.99999, lambda=expected),
                   lower=qpois(0.00001, lambda=expected))
extremes = subset(states, (count > 1.5*upper) | (count < 0.5*lower))
```


### By Zip Codes

```{r}
countByZip = groupBy(both, by=Zip)
zips = join(countByZip, zipPop, by="Zip")
zips = transform(zips, expected=perCapita*Population)
zips = transform(zips, 
                   upper=qpois(0.99999, lambda=expected),
                   lower=qpois(0.00001, lambda=expected))
extremes = subset(zips, (count > 1.5*upper) | (count < 0.5*lower))
```

## Finishing Up

As a **class group** 
* Look for financial or other information available on a by-Zip-code basis.
* Decide what variables you think are important.
* Make a Google Spreadsheet that incorporates these variables.  
* Use the supervised partitioning methods too model how the variables might relate to whether a Zip-code has more or fewer complaints than expected.

